---
title: "Vignette_02_Using_paRsynth_Functions"
author: "GAJ"
date: "2024-12-02"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

```

<h2>How to Use paRsynth Functions</h2>

In this vignette, we will go through how to create vocalizations strings, convert them into Parsons code, calculate frequency anchors, and generate synthetic audio files using the paRsynth package for downstream bioacoustics analyses. 

**Preliminary Step**

Before using the functions in the paRsynth package, there are some preliminary steps we must go through to ensure it will work appropriately and organized neatly.

<u>Step i. Load packages and set paths</u>

It is good practice to clean your global environment before using functions from open-sourced packages. In order for paRsynth functions to run, they need specific external R packages as well. For example, 'write_audio' generates and saves audio files in .WAV format which requires `soundgen::soundgen()` to be installed and loaded. 

```{r message = FALSE, warning = FALSE}

# Clean the global environment
rm(list = ls())

# Specify the required packages
X <- c("devtools", "dplyr", "stringdist", "tidyverse", "ggplot2", "apcluster", "soundgen", "parallel", "stringr", "data.table", "tuneR", "pbapply", "warbleR", "magrittr")

# Install the packages in X if not already installed
is_installed <- function(p) is.element(p, installed.packages()[,1])

invisible(lapply(1:length(X), function(x){
  if(!is_installed(X[x])){
    install.packages(X[x], repos = "http://lib.stat.cmu.edu/R/CRAN")
  }
}))

# Install the paRsynth package from GitHub if you haven't installed it already
# devtools::install_github("gsvidaurre/paRsynth")

# Add paRsynth to the list of packages to load
# X <- c(X, "paRsynth")

# Change "Desktop/.../GitHub_repos" based on where paRsynth is stored on your local machine
testing_path <- "~/Desktop/BIRDS/GitHub_repos/paRsynth/R"

# Load the paRsynth functions that will be tested below
source(file.path(testing_path, "generate_strings.R"))
source(file.path(testing_path, "parsons_code.R"))
source(file.path(testing_path, "frequency_anchors.R"))
source(file.path(testing_path, "write_audio.R"))

# Load all of the packages specified above
invisible(lapply(X, library, character.only = TRUE))

```

<u> Step ii. Initialize working directories for data on your local machine </u>

It is important to have your files organized especially when creating many different types of output data. Therefore, initializing working directories in the beginning will be helpful knowing and understanding the data you create with paRsynth.

```{r eval = TRUE}

# Initialize a base path (this will need to be different per user)
# path <- "/Users/raneemsamman/Desktop" # Raneem's path
# path <- "/Users/gracesmith-vidaurre/Desktop" # Grace's path
path <- "/Users/gretheljuarez/Desktop/BIRDS/paRsynth_Data" # Alexandra's path

# Initialize the directory for analysis on your local computer
analysis_dir <- "paRsynth_methods_synthetic_dataset"

# Combine the base path and the data directory into a single path
analysis_path <- file.path(path, analysis_dir)

# Create the data directory if it doesn't already exist on your computer
if(!dir.exists(analysis_path)){ 
  dir.create(analysis_path)
}

# Specify a folder inside the analysis directory where audio will be written out/read in
audio_dir <- "audio"

# Combine the base path, the analysis directory, and the audio directory into a single path
audio_path <- file.path(path, analysis_dir, audio_dir)

# Create the audio directory if it doesn't already exist on your computer
if(!dir.exists(audio_path)){ 
  dir.create(audio_path)
}

# Specify a folder inside the analysis directory where images will be written out/read in
images_dir <- "images"

# Combine the base path, the analysis directory, and the data directory into a single path
images_path <- file.path(path, analysis_dir, images_dir)

# Create the data directory if it doesn't already exist on your computer
if(!dir.exists(images_path)){ 
  dir.create(images_path)
}

```

<u> Step iii. Check out documentation for the paRsynth functions. These functions are written below in the order in which they should be used </u>

Before using the paRsynth package, it would be beneficial for you to go over the description, usage, arguments, extra details, and examples of each function. 

```{r}

# ?paRsynth::generate_strings
# ?paRsynth::parsons_code
# ?paRsynth::frequency_anchors
# ?paRsynth::write_audio

```

**paRsynth Functions**

<u> STEP 1: Create character strings representing vocal signals with more group membership information than individual information </u>

After installing paRsynth and loading the relevant libraries, you will be able to take the first step into creating synthetic bird calls. The first step is generating the character strings that represent vocalizations using the function generate_strings(). These strings encode both group membership (characters that are shared across individuals within a group) and individual identity (characters that are unique to each individual in a group). The function allows you to specify the length of the string as well as how much of the string encodes group versus individual identity information. It also allows you to control the number of groups, individuals in each group and calls per individual. These variables that you can control are known as the parameters of the function. 

```{r message = FALSE, warning = FALSE}

# Set the seed for reproducibility
set.seed(123)

# Initialize the values of group_information and individual_information, which correspond to the length of the string that will be allocated to encoding group membership or individual identity information
# Group information can outweigh individual information by a factor of at least 10 or higher
factor_ratio <- 10
individual_information <- 2
group_information <- individual_information * factor_ratio
# Initialize the values of the number of groups, individuals, calls, and string length
n_groups <- 2
n_individuals <- 4
n_calls <- 3
string_length <- (group_information + individual_information) + 10

### Generate the string for each call
calls <- generate_strings(
  n_groups = n_groups, 
  n_individuals = n_individuals, 
  n_calls = n_calls, 
  string_length = string_length, 
  group_information = group_information, 
  individual_information = individual_information
  )

dplyr::glimpse(calls)

```

The above chunk of code will generate a data frame called example_calls. Each row of the data frame represents a call, including the group, individual, and call identifiers, along with the vocalization string itself. 

<u> STEP 2: Convert strings to Parsons code </u>

Now that you have a data frame of generated calls, you can use the parsons_code() function to convert the vocalization strings in the data frame example_calls into Parsons code. The goal of this step is to create a representation of the direction of frequency shifts ("up", "down", "constant") that will be used to modulate the sound's pitch when generating synthetic audio later on. This function has 3 parameters: the data frame (df), the name of the column containing the character strings (string_col), and the list that maps each character ("A", "B", "C") to a direction in Parsons code ("up", "down", "constant") (mapping).

```{r}

calls_parsons <- parsons_code(
  df = calls, 
  string_col = "Call", 
  mapping = list("A" = "up", "B" = "down", "C" = "constant")
)

glimpse(calls_parsons)

```

This will return a data frame with all the previous columns in example_calls in addition to one extra column called "Parsons_code", which contains the Parsons code representation of each segment of the vocalization string.

<u> STEP 3: Convert the Parsons code to frequency values </u>

With the Parsons code in your data frame, you are now able to create the frequency anchors that will guide the changes in pitch when creating synthetic audio files using the frequency_anchors() function. It does so by calculating the frequency values for each vocalisation based on the Parsons code from the previous step, a starting frequency value, and a frequency shift value you provide. Therefore, the parameters of this function are the data frame with the parsons code, the names of the columns containing the parsons code (parsons_col), the numeric group identifiers (group_id_col), the numeric individual identifiers (individual_id_col), unique numeric identifiers for each vocalization per individual (call_id_col),  the character string per vocalisation (call_string_col), the baseline frequency value (starting_frequency) which has a default of 4000 Hz, and a value of frequency shift (frequency_shift) which has a default value of 1000. 

```{r}

calls_parsons_frequencies <- frequency_anchors(
  df = calls_parsons, 
  parsons_col = "Parsons_Code", 
  group_id_col = "Group", 
  individual_id_col = "Individual", 
  call_id_col = "Call_ID", 
  call_string_col = "Call", 
  starting_frequency = 4000, 
  frequency_shift = 1000
  )

glimpse(calls_parsons_frequencies)

```

This will return an updated data frame that includes all the previous columns in addition to additional columns that represent the frequency anchors for each vocalization. 

<u> STEP 4: For each call or string, use these frequency vectors to generate synthetic audio files with the soundgen package </u>

Now, finally, you are equipped with all the necessary data to generate the audio files that will allow you to hear the synthetic bird calls we generated! To do that, we will use the function write_audio(). This function creates a sound file (in .wav format) for each vocalization and saves it to your desired directory. Each file will represent one vocalization with its own unique frequency modulation pattern that was developed in step 3. This function has five parameters: the data frame with frequency anchors and string metadata (df), the directory where the audio files will be saved (save_path), the sampling rate for the audio files with a default of 150000 Hz (sampling_rate), the length of the syllable in milliseconds with a default of 200 ms (sylLen), and the prefix for the audio file names (prefix). Note that the prefix you use for the synthetic audio files should match the social level that should contain the most information (here it is the group level because the group information string is longer).

```{r}

synthetic_call_metadata <- write_audio(
  df = calls_parsons_frequencies, 
  save_path = audio_path, 
  sampling_rate = 150000, 
  sylLen = 400, 
  prefix = "GroupMembership"
  )

glimpse(synthetic_call_metadata)

```

This will create files with the extension .wav for each vocalization and will save them in your desired and specified directory with a prefix of your preference.

<u> Step 5: Save the metadata for the synthetic audio files to a CSV file </u>

The data frames that you generate in R using paRsynth can be saved as physical files in your working directory. You can also save your paRsynth generated metadata into a CSV file. I recommend using this file type since it is compatible with R, Microsoft Word, and other software. You can use the function `write.csv()` to save data frames to .csv spreadsheets on your computer:

```{r}

# Add extra metadata for upcoming image files names and rearrange columns
synthetic_call_metadata %>%
  write.csv(., file = file.path(analysis_path, "synthetic_call_metadata.csv"), row.names = FALSE) 

```

Now that you have generated your metadata with paRsyth and it has been saved as a CSV file, navigate to Vignette 03 Analysis to learn how to use the generated outputs of the package described above for downstream bioacoustics analysis.

**2nd Example - Individual Information > Group Information**

Let us do a 2nd example where individual information is greater than group information. Run the same steps above, but interchange 'individual_information' and 'group_information in the code. Make sure to reassign group_information to an even integer larger than 8 and reassign factor_ratio to a smaller even integer starting from 2 - 10.

```{r eval = TRUE}

# Set the seed for reproducibility
set.seed(456)

# Interchange individual_information with group_information. Reassign factor_ratio to 2 and group_information to 8
factor_ratio2 <- 2
group_information2 <- 8
individual_information2 <- group_information2 * factor_ratio2
n_groups2 <- 2
n_individuals2 <- 4
n_calls2 <- 3
string_length2 <- (group_information + individual_information) + 10

### Generate the string for each call
calls2 <- generate_strings(
  n_groups = n_groups2, 
  n_individuals = n_individuals2, 
  n_calls = n_calls2, 
  string_length = string_length2, 
  group_information = group_information2, 
  individual_information = individual_information2
  )

### Convert each string to Parsons code
calls_parsons2 <- parsons_code(
  df = calls2, 
  string_col = "Call", 
  mapping = list("A" = "up", "B" = "down", "C" = "constant")
)

### Convert each Parsons code to frequency values
calls_parsons_frequencies2 <- frequency_anchors(
  df = calls_parsons2, 
  parsons_col = "Parsons_Code", 
  group_id_col = "Group", 
  individual_id_col = "Individual", 
  call_id_col = "Call_ID", 
  call_string_col = "Call", 
  starting_frequency = 4000, 
  frequency_shift = 1000
  )

### For each call, use the frequency vectors to generate synthetic audio files with the soundgen package
synthetic_call_metadata2 <- write_audio(
  df = calls_parsons_frequencies2, 
  save_path = audio_path, 
  sampling_rate = 150000, 
  sylLen = 400, 
  prefix = "IndividualIdentity"
  )

### Save the metadata for the synthetic audio files to a CSV file
# Add extra metadata for upcoming image files names and rearrange columns
synthetic_call_metadata2 %>%
  write.csv(., file = file.path(analysis_path, "synthetic_call_metadata2.csv"), row.names = FALSE)

glimpse(synthetic_call_metadata2)
```

After running this code you will have two datasets: 1) reflects vocalizations with more group information than individual and 2) vice versa. This will be extremely beneficial when conducting downstream bioacoustic analysis. Navigate to Vignette 03 to learn how to analyze these outputs you created with paRsynth.
