---
title: "Vignette_02_How_Functions_Work"
author: "GAJ"
date: "2024-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<h2>Control information encoding in synthetic vocal signals</h2>

<u>STEP 1: Generate vocalizations strings using generate_strings()</u>

After installing paRsynth and loading the relevant libraries, you will be able to take the first step into creating synthetic bird calls. The first step is generating the character strings that represent vocalizations using the function generate_strings(). These strings encode both group membership (characters that are shared across individuals within a group) and individual identity (characters that are unique to each individual in a group). The function allows you to specify the length of the string as well as how much of the string encodes group versus individual identity information. It also allows you to control the number of groups, individuals in each group and calls per individual. These variables that you can control are known as the parameters of the function. 

```{r}
source("~/Desktop/BIRDS/GitHub_repos/paRsynth/R/generate_strings.R")
source("~/Desktop/BIRDS/GitHub_repos/paRsynth/R/parsons_code.R")
source("~/Desktop/BIRDS/GitHub_repos/paRsynth/R/frequency_anchors.R")
source("~/Desktop/BIRDS/GitHub_repos/paRsynth/R/write_audio.R")

library(tidyverse)
library(soundgen)

# A random seed for reproducibility
set.seed(8)

# Generating an example calls with 2 groups, 5 individuals per group, and 10 calls per individual with group information (8 characters) being econded 4 times more than the individual information (2 characters)
example_calls <- generate_strings(n_groups = 2, n_individuals = 5, n_calls = 10, string_length = 16, group_information = 8, individual_information = 2)

# to view the generated calls
glimpse(example_calls)
```

The above chunk of code will generate a data frame called example_calls. Each row of the data frame represents a call, including the group, individual, and call identifiers, along with the vocalization string itself. 

<u>STEP 2: Convert strings to Parsons code</u>

Now that you have a data frame of generated calls, you can use the parsons_code() function to convert the vocalisations strings in the data frame example_calls into Parsons code. The goal of this step is to create a representation of the direction of frequency shifts ("up", "down", "constant") that will be used to modulate the sound's pitch when generating synthetic audio later on. This function has 3 parameters: the data frame (df), the name of the column containing the character strings (string_col), and the list that maps each character ("A", "B", "C") to a direction in Parsons code ("up", "down", "constant") (mapping).

```{r}
# converting the generated vocalization strings in example_calls df to Parsons code
example_calls_parsons <- parsons_code(example_calls, "Call", list("A" = "up", "B" = "down", "C" = "constant"))

# to view the resulted parsons code 
glimpse(example_calls_parsons)
```

This will return a data frame with all the previous columns in example_calls in addition to one extra column called "Parsons_code", which contains the Parsons code representation of each segment of the vocalisation string.

<u>STEP 3: Generate frequency anchors</u>

With the Parsons code in your data frame, you are now able to create the frequency anchors that will guide the changes in pitch when creating synthetic audio files using the frequency_anchors() function. It does so by calculating the frequency values for each vocalisation based on the Parsons code from the previous step, a starting frequency value, and a frequency shift value you provide. Therefore, the parameters of this function are the data frame with the parsons code, the names of the columns containing the parsons code (parsons_col), the numeric group identifiers (group_id_col), the numeric individual identifiers (individual_id_col), unique numeric identifiers for each vocalization per individual (call_id_col),  the character string per vocalisation (call_string_col), the baseline frequency value (starting_frequency) which has a default of 4000 Hz, and a value of frequency shift (frequency_shift) which has a default value of 1000. 

```{r}
# generating frequency anchors from the Parsons code
anchors <- frequency_anchors(example_calls_parsons, "Parsons_Code", "Group", "Individual", "Call_ID", "Call", starting_frequency = 4000, frequency_shift = 1000)

# to view the result
glimpse(anchors)
```

This will return an updated data frame that includes all the previous columns in addition to additional columns that represent the frequency anchors for each vocalisation. 

<u>STEP 4: Generate and save audio files</u>

Now, finally, you are equipped with all the necessary data to generate the audio files that will allow you to hear the synthetic bird calls we generated! To do that, we will use the function write_audio(). This function creates a sound file (in .wav format) for each vocalisation and saves it to your desired directory. Each file will represent one vocalisation with its own unique frequency modulation pattern that was developed in step 3. This function has five parameters: the data frame with frequency anchors and string metadata (df), the directory where the audio files will be saved (save_path), the sampling rate for the audio files with a default of 150000 Hz (sampling_rate), the length of the syllable in milliseconds with a default of 200 ms (sylLen), and the prefix for the audio file names (prefix). Note that the prefix you use for the synthetic audio files should match the social level that should contain the most information (here it is the group level because the group information string is longer).

```{r}
# setting the path to save the audio files
 path <- "~/Desktop" # Update this path to reflect your own directory structure
 tmp_dir <- "testing"
 tmp_path <- file.path(path, tmp_dir)

 # Create the temporary directory if it doesn't already exist
 if(!dir.exists(tmp_path)){
  dir.create(tmp_path)
 }

# generate and save audio files for a subset of calls
set.seed(8)

inds <- sample(1:nrow(anchors), 10, replace = FALSE)
write_audio(anchors[inds, ], save_path = tmp_path, sampling_rate = 150000, sylLen = 200, prefix = "GroupSignatures")

# Remove the temporary directory and all files within it
if(tmp_path == file.path(path, tmp_dir)){
  unlink(tmp_path, recursive = TRUE)
 }
```

This will create files with the extension .wav for each vocalizations and will save them in your desired and specified directory with a prefix of your preference.

<u>STEP 5: Generate Vocalizations with more Individual Identity Information and save audio files</u>

Run the same steps above, but instead create a data set of synthetic vocalizations with more individual information than group information (This will require changes to the arguments in generate_strings(), as well as the prefix used in write_audio().)
